---
title: "District-Locations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Importing Packages

```{r, message=F}
library(tidyverse)
library(readxl)
library(corrplot)
library(tigris)
```

# Importing and Cleaning Data

```{r}
district_data <- read_excel("C:/Users/User/Documents/GitHub/household-conditions/data/NHGIS_District_data.xslx")
View(district_data)
```
names_list <- c(
    "school_ID", 
    "state", 
    "dist",  # Geographic School District 
    "children", # Children 5-17 (SAIPE Estimate)
    "pct_pov", # % Poverty (SAIPE Estimate)
    "pct_SP", # % Single Parent Estimate
    "SP_MOE", # Single Parent Margin of Error
    "pct_HHVJ", # % HHs With Vulnerable Job Estimate
    "HHVJ_MOE", # Vulnerable Job Margin of Error
    "pct_CC", # % Crowded Conditions Estimate
    "CC_MOE", # HH With Crowded Conditions Margin of Error
    "pct_NCI", # % No Computer or Internet Estimate
    "nci_MOE", # No Computer or Internet Margin of Error
    "pct_CD", # % Children with Disability
    "CD_MOE", # Children with Disability Margin of Error
    "pct_CLI", # % Linguistically Isolated Children
    "CLI_MOE" # Linguistically Isolated Children Margin of Error
)
names(district_data) <- names_list
district_pcts <- district_data %>%
    select(-ends_with("MOE")) 
View(district_data)

```

# Need to separate the original data by year's to be able to merge with district location data

```{r}




```


