---
title: "Joining Race Data with District Data"
author: "Noel Goodwin, Abigail Joppa, Jon Geiger"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(ggthemes)
theme_set(theme_clean())
```

### Import data

```{r, message=FALSE}
race_data <- read_delim("../raw/race_raw.txt", delim = "|", escape_double = FALSE, trim_ws = TRUE)
hh <- read_csv("../data/hh.csv")
```



```{r}
names(race_data)
```

### Renaming Variables

I used the data dictionary provided to rename the variable names. 
*note: It looks like they only looked combined all of the Asian population which is a pretty big umbrella. It may be worth looking into what exactly this variable includes. 
```{r}
race_names<- c("geo_id",
              "dist", 
              "leaid", 
              "year", 
              "Iteration",
               "total_pop_est", #Total population
               "total_pop_moe", # total population MOE
               "total_hisp_latino", #Total population; Hispanic or Latino (of any race)
               "total_hisp_latino_moe", #Total population; Hispanic or Latino (of any race) MOE
               "pct_hisp_latino", # % total population: Hispanic or Latino (of any race)
               "pct_hisp_latino_moe", # % total population: Hispanic or Latino (of any race) MOE
               "total_mexican", # total mexican population
               "total_mexican_moe", # total mexican population MOE
               "pct_mexican", # % mexican population
               "pct_mexican_moe", # % mexican population MOE
               "total_puertrican", # total puertorican population
               "total_puertrican_moe", # total puertorican population MOE
               "pct_puertrican",  # % puertorican population
               "pct_puertrican_moe", # % puertorican population MOE
               "total_cuban", # total cuban population
               "total_cuban_moe", # total cuban population MOE
               "pct_cuban", # % cuban population
               "pct_cuban_moe", # % cuban population MOE
               "total_other_hl", # total population of other hispanic and latino
               "total_other_hl_moe", # total population of other hispanic and latino MOE
               "pct_other_hl", # % population of other hispanic and latino
               "pct_other_hl_moe", # % population of other hispanic and latino MOE
               "total_NOT_hl", # total population that are not hispanic nor latino
               "total_NOT_hlmoe", # total population that are not hispanic nor latino MOE
               "pct_NOT_hl", # % population that are not hispanic nor latino
               "pct_NOT_hl_moe", # % population that are not hispanic nor latino MOE
               "total_white", #total population white
               "total_white_moe", #total population white MOE
               "pct_white", # % population white
               "pct_white_moe", # % population white MOE
               "total_black", #total population black or african american
               "total_black_moe", #total population black or african american MOE
               "pct_black", # % population black or african american
               "pct_black_moe", # % population black or african american MOE
               "total_native", #total native population 
               "total_native_moe", #total native population MOE
               "pct_native", # % native population
               "pct_native_moe", # % native population MOE
               "total_asian", #total asian population 
               "total_asian_moe", #total asian population MOE
               "pct_asian", # % asian population
               "pct_asian_moe", # % asian population MOE
               "total_PI", # total PI = Native Hawiian/PAcific Islander 
               "total_PI_moe", # total PI = Native Hawiian/PAcific Islander MOE
               "pct_PI", # % PI 
               "pct_PI_moe", # % PI MOE
               "total_other", #total people of other race (nonhisp/latino)
               "total_other_moe", #total people of other race (nonhisp/latino) MOE
               "pct_other", # % people of other race (nonhisp/latino)
               "pct_other_moe", #% people of other race (nonhisp/latino) MOE
               "total_nonhl_2race", #non hispanic or latio with 2 races or more
               "total_nonhl_2race_moe", #non hispanic or latio with 2 races or more MOE
               "pct_nonhl_2race", # % non hispanic or latio with 2 races or more
               "pct_nonhl_2race_moe", # % non hispanic or latio with 2 races or more MOE
               "total_nonhl_2_other", #Total population; Not Hispanic or Latino; Two or more                                          races; Two races including Some other race
               "total_nonhl_2_other_moe", #MOE
               "pct_nonhl_2_other", # % population; Not Hispanic or Latino; Two or more                                          races; Two races including Some other race
               "pct_nonhl_2_other_moe", 
               "total_nonhl_2_3other", # Total population; Not Hispanic or Latino; Two or more                                          races; Two races excluding Some other race, and Three                                           or more races
               "total_nonhl_2_3other_moe", 
               "pct_nonhl_2_3other", # % population; Not Hispanic or Latino; Two or more                                               races; Two races excluding Some other race, and Three                                           or more races
               "pct_nonhl_2_3other_moe"
)
names(race_data) <- race_names
race_data <- race_data %>%
  select(-"Iteration") %>% # This variable is not relevant, as it is the same through each entry. We can keep in mind that it is 105.
  mutate(leaid = as.integer(leaid))
  
race_data %>% head(10)
```

### What is a Geo ID? 
It appears that the geoid** is a combination of some sort of US code (97000US), followed by a two digit state fips code, and lastly the district ID used in the original NHGIS district data. The same names of school districts, without specification about which state they are from (see Lincoln County School District). This may be problematic in the joining process. 

```{r}
race_data %>%
filter(str_detect(dist, 'Lincoln County School District'))
```

Update: After some exploration, we now know that that the two digit state code (fips) and the district ID make up the leaid which is already included in the race data. Instead of separating the geoid as we did previously, we will use the leaid to join with the household data. 

### Joining using leaid

There are 1,404 missing values which is the difference between the two data sets - household data (13314), race_data (11910). This is because the household data is missing 1404 leaids. We successfully joined all of the available race data with the district data. 
```{r}

dist_race_join <- hh %>%
  left_join(race_data, by=c("leaid"))%>%
  select(-dist.y)%>%
  rename(dist = dist.x) %>%
  relocate(where(is.numeric), .after = where(is.character))


colSums(is.na(dist_race_join))


```

### Missing school districts

Lets see which districts did not join and therefore are the missing school districts in the race data set. 

```{r}

missing_race_dist <- dist_race_join %>% 
  filter_all(any_vars(is.na(.))) %>%
  select(leaid, state, dist, children)

missing_race_dist %>% head(10)

```

This may not be too big of an issue if most of those schools are small districts that we were going to drop anyway, but we can see that there are about 463 missing school districts that have more than 150 students which may be important for our analysis. 
```{r}
missing_race_dist %>%
  filter(children >= 150)%>%
  dim()
```

### Next steps

- Figure out why there are 1,404 missing school districts in the race data and see if the missing data is accessible elsewhere

